---
title: "Week 6: Lab"
subtitle: "Sequence alignment and phylogenetics"
format: pdf
editor: visual
date: now
---

## Load packages

```{r}
#| warning: false
#| message: false
#| output: false

pkgs <- c("ggplot2","devtools","adegenet","phangorn", "ape", "ade4",
          "Biostrings", "DECIPHER", "ggtree", "pwalign")

for (i in seq_along(pkgs)){
  pkg <- pkgs[[i]]
  tryCatch(if(pkg%in%rownames(installed.packages())){
    cat("Available:", pkg, "\n")
    library(pkg, character.only = TRUE)
    cat("Package loaded:", pkg,"\n")
  }
  else if(!(pkg%in%rownames(installed.packages()))){
    install.packages("BiocManager")
    BiocManager::install(pkg)
    library(pkg, character.only = TRUE)
    cat("Package loaded:", pkg,"\n")
  }
 else{
    cat("Not available:",pkg,">> now installing...", "\n")
    install.packages(pkg, repos =  "https://cloud.r-project.org")
    cat("Now available:", pkg, "\n")
    library(pkg, character.only = TRUE)
    cat("Package loaded:", pkg,"\n")
  }
  )
}

```

```{r}
#| eval: false

##############################################################################
## After running the above, if you cannot find any of the above packages in R,
## or it's giving you an erro, then, run them individual as below
##############################################################################

###############################################################################
## For "ggplot2","devtools","adegenet","phangorn", "ape", "ade4", "BiocManager"
###############################################################################
install.packages("ggplot2") # do that for all
library("ggplot2") #load it

################################################################
## For "Biostrings", "DECIPHER", "ggtree", "pwalign"
###############################################################
BiocManager::install("Biostrings") # do that for all
library(Biostrings) #load it


```

## 1) Sequence Alignment

### a) Pairwise alighment

```{r}
#| output: false
#| eval: false

## These are the sequences we are working with for pairwise alignment

## Query sequence
Seqq <- "ATTTGAATGGATGTCAATCCGACTCTACTGTTCCTAAAGGTTCCAGCGCAAAATGCCATAAGCACCACATT
CCCTTATACTGGAGATCCTCCATACAGCCATGGAACAGGAACAGGGTACACCATGGACACAGTCAACAGAACACACCAAT
ATTCAGAGAAGGGGAAGTGGACGACAAATACAGAAACTGGGGCACCCCAACTCAACCCAATTGATGGACCACTACCTGAG
GATAATGAGCCAAGTGGATATGCACAAACAGACTGTGTCCTGGAGGCTATGGCCTTCCTTGAAGAATCCCACCCAGGTAT
CTTTGAGAACTCATGCCTTGAAACAATGGAAGTCGTTCAACAAACAAGGGTGGACAAACTAACCCAAGGTCGCCAGACTT
ATGATTGGACATTAAACAGAAATCAACCGGCAGCAACTGCATTAGCCAACACCATAGAAGTTTTTAGATCGAATGGACTA
ACAGCTAATGAATCGGGAA" #GenBank: CY002062.1

## Query sequence 2
Seqq2 <- "AGCAAAAGCAGGGGATAATTCTATTAACCATGAAGACTATCATTGCTTTGAGCTACATTCTATGTCTGGT
TTTCGCTCAAAAACTTCCCGGAAATGACAACAGCACGGCAACGCTGTGCCTTGGGCACCATGCAGTACCAAACGGAACGA
TAGTGAAAACAATCACGAATGACCAAATTGAAGTCACTAATGCTACTGAACTGGTTCAGAGTTCCTCAACAGGTGGAATA
TGCGACAGTCCTCATCAGATCCTTGATGGAGAAAACTGCACACTAATAGATGCTCTATTGGGAGACCCTCAGTGTGATGG
CTTCCAAAATAAGAAATGGGACCTTTTTGTTGAACGCAGCAAAGCCTACAGCAACTGTTACCCTTATGATGTGCCGGATT
ATGCCTCCCTTAGGTCACTAGTTGCCTCATCCGGCACACTGGAGTTTAACAATGAAAGCTTCAATTGGACTGGAGTCACT
CAAAATGGAACAAGCTCTGC" #GenBank: CY002064.1

## Reference sequence
Reff <- "ATGAAGACTATCATTGCTTTGAGCTACATTCTATGTCTGGTTTTCGCTCAAAAACTTCCCGGAAATGACAA
CAGCACGGCAACGCTGTGCCTTGGGCACCATGCAGTACCAAACGGAACGATAGTGAAAACAATCACGAATGACCAAATTG
AAGTCACTAATGCTACTGAACTGGTTCAGAGTTCCTCAACAGGTGGAATATGCGACAGTCCTCATCAGATCCTTGATGGA
GAAAACTGCACACTAATAGATGCTCTATTGGGAGACCCTCAGTGTGATGGCTTCCAAAATAAGAAATGGGACCTTTTTGT
TGAACGCAGCAAAGCCTACAGCAACTGTTACCCTTATGATGTGCCGGATTATGCCTCCCTTAGGTCACTAGTTGCCTCAT
CCGGCACACTGGAGTTTAACAATGAAAGCTTCAATTGGACTGGAGTCACTCAAAATGGAACAAGCTCTGCTTGCAAAAGG
AGATCTAATAACAGTTTCT" #NCBI Reference Sequence: NC_007366.1

```


```{r}

Seq1 <- DNAString("ATTTGAATGGATGTCAATCCGACTCTACTGTTCCTAAAGGTTCCAGCGCAAAATGCCATAAGCACCACATTCCCTTATACTGGAGATCCTCCATACAGCCATGGAACAGGAACAGGGTACACCATGGACACAGTCAACAGAACACACCAATATTCAGAGAAGGGGAAGTGGACGACAAATACAGAAACTGGGGCACCCCAACTCAACCCAATTGATGGACCACTACCTGAGGATAATGAGCCAAGTGGATATGCACAAACAGACTGTGTCCTGGAGGCTATGGCCTTCCTTGAAGAATCCCACCCAGGTATCTTTGAGAACTCATGCCTTGAAACAATGGAAGTCGTTCAACAAACAAGGGTGGACAAACTAACCCAAGGTCGCCAGACTTATGATTGGACATTAAACAGAAATCAACCGGCAGCAACTGCATTAGCCAACACCATAGAAGTTTTTAGATCGAATGGACTAACAGCTAATGAATCGGGAA")

Seq2 <- DNAString("AGCAAAAGCAGGGGATAATTCTATTAACCATGAAGACTATCATTGCTTTGAGCTACATTCTATGTCTGGTTTTCGCTCAAAAACTTCCCGGAAATGACAACAGCACGGCAACGCTGTGCCTTGGGCACCATGCAGTACCAAACGGAACGATAGTGAAAACAATCACGAATGACCAAATTGAAGTCACTAATGCTACTGAACTGGTTCAGAGTTCCTCAACAGGTGGAATATGCGACAGTCCTCATCAGATCCTTGATGGAGAAAACTGCACACTAATAGATGCTCTATTGGGAGACCCTCAGTGTGATGGCTTCCAAAATAAGAAATGGGACCTTTTTGTTGAACGCAGCAAAGCCTACAGCAACTGTTACCCTTATGATGTGCCGGATTATGCCTCCCTTAGGTCACTAGTTGCCTCATCCGGCACACTGGAGTTTAACAATGAAAGCTTCAATTGGACTGGAGTCACTCAAAATGGAACAAGCTCTGC")

Ref <- DNAString("ATGAAGACTATCATTGCTTTGAGCTACATTCTATGTCTGGTTTTCGCTCAAAAACTTCCCGGAAATGACAACAGCACGGCAACGCTGTGCCTTGGGCACCATGCAGTACCAAACGGAACGATAGTGAAAACAATCACGAATGACCAAATTGAAGTCACTAATGCTACTGAACTGGTTCAGAGTTCCTCAACAGGTGGAATATGCGACAGTCCTCATCAGATCCTTGATGGAGAAAACTGCACACTAATAGATGCTCTATTGGGAGACCCTCAGTGTGATGGCTTCCAAAATAAGAAATGGGACCTTTTTGTTGAACGCAGCAAAGCCTACAGCAACTGTTACCCTTATGATGTGCCGGATTATGCCTCCCTTAGGTCACTAGTTGCCTCATCCGGCACACTGGAGTTTAACAATGAAAGCTTCAATTGGACTGGAGTCACTCAAAATGGAACAAGCTCTGCTTGCAAAAGGAGATCTAATAACAGTTTCT")

pwalign1 <- pairwiseAlignment(Seq1, Ref)
pwalign1
```


### b) Multiple alignment


```{r}
#| warning: false

Mseq <- list()
  
Mseq[1] <-  Seq2
Mseq[2] <- Seq1
Mseq[3] <- Ref

dna0 <- DNAStringSet(Mseq)
Malign <- AlignSeqs(dna0)
Malign
```

\newpage

## 2) Phylogenetic analysis

```{r}

dna_df <- read.dna("http://adegenet.r-forge.r-project.org/files/usflu.fasta",
                   format = "fasta")
###############################################################################
## If dna_df is not loading, download GENET_fasta.RData from the Week 6 folder
## Copy the path from your PC and paste in the load() below. It will be loaded to R
## E.g, load("C:/Users/Kyles/Download/GENET_fasta.RData")
###############################################################################

# load("")

## Perform Distance between sequence
dna_dist <- dist.dna(dna_df, model = "TN93")

dna_tree <- as.phylo(hclust(dna_dist, method = "average"))

```

### a) Model selection

```{r}

To_phyD <- phyDat(dna_df)

mod.phy <- modelTest(To_phyD, model = c("JC", "F81", "K80", "HKY","K81",
                             "SYM", "GTR"))
```

### b) Plot trees

```{r}
#| warning: false

## View node numbering
ggtree(dna_tree)+
  geom_tiplab()+
  geom_text2(mapping = aes(label = node))

## Add tip label layer to the tree
gtree <- ggtree(dna_tree)+
  geom_tiplab()

## View subset of the node
viewClade(gtree, node = 83) 

## Tree layouts

ggtree(dna_tree, layout = "radial")+
  geom_tiplab()

ggtree(dna_tree, layout = "fan")+
  geom_tiplab()
```

### c) Using Neighor-joining

```{r}
#| warning: false
#| message: false

## Compute tree using neighbor-joining
nj_tree <- nj(dna_dist)

## Convert to phylo object
nj_tree2 <- as.phylo(nj_tree)

## Plot tree
nj_gtree <- ggtree(nj_tree2)+
  geom_tiplab()

## View a subset node
viewClade(nj_gtree, node=83)
```

### d) UPGMA

```{r}
#| warning: false
#| message: false

## Compute tree using the unweighted pair group method with arithmetic mean (UPGMA)
upgma_tree <- upgma(dna_dist)

## Convert to phylo object
upgma_tree2 <- as.phylo(upgma_tree)

## Plot tree
upgma_gtree <- ggtree(upgma_tree2)+
  geom_tiplab()

## View a subset node
viewClade(upgma_gtree, node=83)
```

## Task

-   Replace TN93 with the best model (lowest AIC or BIC) in 2a above. That is, choose any of "JC", "K80", "F81", or "K81" model that has the lowest value.

-   Compare the tree with 2c and 2d

```{r}
#| warning: false
#| message: false

## Perform Distance between sequence
dna_mat <- dist.dna(dna_df, model = "K81") #Replace TN93

###############################################
### NJ #####################################
###############################################

nj_tr <- as.phylo(nj(dna_mat))

#Plot tree
nj_gtree <- ggtree(nj_tree2)+
  geom_tiplab()

#View a subset node
viewClade(nj_gtree, node=83)

###############################################
### UPGMA #####################################
###############################################

upgma_tr <- as.phylo(upgma(dna_mat, method = "average"))

## Plot tree
upgma_gtree <- ggtree(upgma_tree2)+
  geom_tiplab()

## View a subset node
viewClade(upgma_gtree, node=83)

```

Is there any difference?
There's no difference between the task method and 2c&d

Note:

Unlike the UPGMA, the neighbor-joining method does not make all taxa equidistant from the root. This is a crucial difference as UPGMA makes the molecular clock assumption (i.e, uniform evolutionary rates) which NJ doesn't make. This is evident as you'll notice equidistant tips in UPGMA compared to different tip length in NJ both in the task and 2c&d.
